---
title: "Single-species models in discrete time"
subtitle: "Problem Set 1"
author: "Thomas O'Leary"
date: "September 11, 2019"
geometry: "left = 2cm, right = 2cm, top = 1.5cm, bottom = 1.5cm"
output: 
  pdf_document:
    fig_width: 6
    fig_height: 3.75
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section{1. Discrete-time logistic growth model: Increasing r}
Modify the following code to determine the long-term behavior of the discrete-time logistic growth model. In the example below, the population movestowards an equilbrium point. How does the long-term behavior change as you increase $r$ slowly.

```{r, echo = TRUE, eval = TRUE}
plot_log_growth <- function(r, max_time = 50, K = 1, FUN = points, 
                            color = "black"){
  # initialize vector
  N <- vector('numeric',length = max_time)
  N[1] <- 0.1
  # iterate model
  for (t in 1:max_time){
    N[t+1] <- N[t] + r*N[t]*(1 - N[t]/K)
  }
  # plot N over time
  FUN(N, type = 'l', ylim = c(0, 1.25), col = color, 
      main = "Logistic growth at increasing values of r", xlab = "time (t)",
      ylab = "N (population size)")
}
# plot a bunch of different growth curves at increasing values of r
plot_log_growth(r = 0.4, FUN = plot)
plot_log_growth(r = 1, color = "red")
plot_log_growth(r = 1.4, color = "orange")
plot_log_growth(r = 1.8, color = "yellow")
plot_log_growth(r = 2.0, color = "green")
plot_log_growth(r = 2.5, color = "blue")
```


As $r$ increases the amount of fluctuation at the equilibrium point increases. So at $r = 0.4$ $N$ gradually increases to 1.0 and then stays at equilbrium. At higher $r$ values the population reaches quilibrium faster, but begins to fluctuate. At values close to 2.0 the equilibrium begins to oscillate above and below an $N$ of 1.0. 


\section{2. Plot the final ten years of N as a function of increasing values of $r$ }
To show the results in question 1 more clearly, write R code to build a plot of the equilibrium value versus the value for $r$. To do this, run the model for 50 years, and only plot the last 10 years. If the solution is an equilibrium point, all 10 years should be the same values.

```{r, echo = TRUE, eval = TRUE}
max_time <- 50
K <- 1

n_vals <- NULL
r_vals <- NULL
for (r in seq(from = 0.4, to = 3, by = .1)){
  # initialize vector
  N <- vector('numeric',length = max_time)
  N[1] <- 0.1
  # iterate model
  for (t in 1:max_time){
    N[t+1] <- N[t] + r*N[t]*(1 - N[t]/K)
  }
  # save only the final 10 years and the corresponding r values
  n_vals <- c(n_vals, N[(max_time-9):max_time])
  r_vals <- c(r_vals, rep(r, 10))
}

vals <- as.data.frame(cbind(r_vals, n_vals))

# load ggplot2 library to run ggplot()
library(ggplot2)

ggplot(vals, aes(x = r_vals, y = n_vals)) +
  geom_jitter(mapping = aes(x = r_vals, y = n_vals), fill = "grey", 
        alpha = 0.5, size = 3, pch = 21, color = "black", width = 0.02) +
  theme_classic() + 
  labs(x = "r value", y = "N (population size)", 
       title = "Population size at \"equilibrium\" as a function of r")
```